Testing Charts: viewer's perceptual accuracy in surveys

### 2023-07-14 SDSS-2023/testing-charts-accuracy.qmd
starting to fill in outline for intro, etc.

---
title: "Testing Charts: viewer's perceptual accuracy in surveys"
short-title: Testing charts in surveys # running title
author:
  - first: Kiegan                 # first name
    first-init: K.              # initial of first name
    last: Rice                  # last name
    marker: 1               # identifier for the affiliation
    footnote-id: 1              # optional identifier for the footnote
    email: foo@bar.com          # optional email address
  - first: Heike
    first-init: H.
    last: Hofmann
    marker: 2
  - first: Nola
    first-init: N.
    last: duToit
    marker: 1
  - first: Edward
    first-init: E.
    last: Mulrow
    marker: 1
footnote:
  - content: Corresponding author
    type: corresp               # set type for the corresponding author
    id: 1
affiliation:
  - institution: National Opinion Research Center (NORC)
    country: United States
    marker: 1
  - institution: Iowa State University
    department: Statistics
    country: United States
    marker: 2
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication, and decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in statistical charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'list, keywords, here, 3-6, are, fine'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
format:
  pdf: default
documentclass: jds
---

```{=latex}
  \newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
  \newcommand{\kr}[1]{{\textcolor{teal}{#1}}}
  \setlength{\parindent}{0pt}
```

\hh{hello?}

@@

<!-- fig-width: 6.5 -->
<!-- fig-height: 4.5 -->
<!-- template: template-jdsart.cls.tex -->

@@

<!-- What is our main objective? -->
What do viewers see, when we show them a data chart? A crucial step in the process of 'understanding' a chart, is that a visual allows us to make comparisons between its parts. @clevelandGraphicalPerceptionTheory1984 defined for the purpose of their seminal study the better visual as the one that allows viewers to make more accurate comparisons. Based on mappings of quantitative variables to different graphical elements, 
Cleveland and McGill's study resulted in a ranking of perceptual tasks from most accurate to least accurate. 
@mackinlayAutomatingDesignGraphical1986 took this approach and expanded the tasks and their order to  ordinal and nominal scales, as shown in @fig-perceptual-tasks.

@@

<!-- Bias from crowd-sourcing? sample size?  -->
Crowd workers are known to be biased towards male, young and relatively higher education: @borgoCrowdsourcingInformationVisualization2017

@@

```{r}
#| label: fig-aligned-unaligned
#| fig-cap: "The only difference between the two pairs of rectangles A, B and C, D is their alignment, i.e. A and C are of identical size, as are B and D. While the predominant response for the unaligned pair on the left is 'they are of the same size', more than half of the viewers respond with 'D is bigger' to the aligned pair of bars on the right."
#| fig-subcap:
#| - "**Unaligned bars**"
#| - "**Aligned bars**"
#| layout-nrow: 1
#| fig-align: "center"
#| out-width: 50%
knitr::include_graphics("images/round5-modified.PNG")
knitr::include_graphics("images/round3-modified.PNG")
```

<!-- motivate the use of just noticeable differences -->
Asking perceptual questions in a survey is different from the controlled environment of a cognitive lab, where these kind of questions would usually be addressed. This means, that instructions to participants have to be delivered in a very short and easily understandable, because questions arising from the task can not be answered. Similarly, rather than asking the same (or similar) type of question with varied signals multiple hundred times, in a survey we can ask only a few questions. In order to observe any effect, we need to ask questions that are perceptually hard, which means that we need to ask questions about stimuli that are close to our perceptual threshold. The **Just-Noticeable Difference** (JND) is defined as the smallest difference that will be detected 50% of the time. We are using results from studies on barcharts and pie charts [@luModelingJustNoticeable2022] to inform the differences in charts shown to survey panelists.

- How do structural design choices in a data visualization impact viewers’ ability to identify the larger of two elements?

- How do aesthetic design choices in a data visualization impact viewers’ ability to identify the larger of two elements?

@@

Aesthetic design choices (structural choices seems stronger/there could be a lot to talk about there… should we skip aesthetic on this one?)

@@

Binary accuracy (correct/incorrect – ‘they are the same’ is incorrect here)

@@

I’ve noted this below, but: how to model? Ordinal response? Binary (certain or very certain vs everybody else)?

@@

<!-- describe stimulus -->
Both charts in \autoref{fig-tasks-12} show the same data with slight modifications to the order of the levels – the first and second level in each of the bars are reversed between the left and the right chart. Participants were asked to compare the relative sizes of the tiles marked A and B (C and D, respectively) and select the correct response out of the possible choices:

@@

<!-- |                                   |                                   | -->
<!-- |-----------------------------------|-----------------------------------| -->
<!-- | **Unaligned positions**           | **Aligned positions**             | -->
<!-- | ![Unaligned positions marked A and B](../SDSS-2023/images/vertical-ab.PNG) | ![Aligned positions marked C and D](../SDSS-2023/images/vertical-cd.PNG) | -->

@@

## Data Analysis and Results

@@

Question for Ed: If we compare a full sample to a split sample, do we still want to weight these ‘equally’?

Analysis should be done using the ‘survey’ package and weights should be taken into account.

@@

The data used for assessing the accuracy of comparisons in \autoref{tbl-tasks-12} is collected in two rounds of the NORC Omnibus survey. Rounds 1 and 2 are combined by adjusting the weights  with $\lambda$ = `r round(lambda[1],3)` for an effective sample size of `r round(neff(round_12$weight),1)`. Figure \ref{fig-alignment}(a) shows that more than twice the number of responses is accurate, when the tiles are aligned along the same axis. 
Because each participant was shown both versions of the chart, we can use a paired $t$-test to compare mean accuracy between the two charts. The resulting $t$-statistic is highly significant ($t$ statistic: `r round(paired$statistic,1)`, df: `r paired$parameter`, $p$-value: < 2.2e-16).

@@

STRUCTURAL VARIATION – START HERE

@@

% yes across each different structural condition

@@

% zoomed by device + alignment (already have this chart)

@@

Model 2 (this may not be feasible for comparison depending on what level the ‘TOTALTIME’ is captured at):

@@

Same models as above, but I’m not sure how we want to do the response. Ordinal response? Binary (certain or very certain vs everybody else)?

AESTHETIC VARIATION – ONLY IF TIME

Binary accuracy (correct/incorrect – ‘they are the same’ is incorrect here) across structural choices

@@

Same models as above, but I’m not sure how we want to do the response. Ordinal response? Binary (certain or very certain vs everybody else)?

@@

# Supplementary Material {-}

+ **Participant Data (Linear):** Link to csv file with the data. 
+ **Data Analysis Code:** Link to an html document with annotated code chunks.

>>>>

---
title: "Testing Charts: viewer's perceptual accuracy in surveys"
short-title: Testing charts in surveys # running title
author:
  - first: Kiegan                 # first name
    first-init: K.              # initial of first name
    last: Rice                  # last name
    marker: 1               # identifier for the affiliation
    footnote-id: 1              # optional identifier for the footnote
    email: foo@bar.com          # optional email address
  - first: Heike
    first-init: H.
    last: Hofmann
    marker: 2
  - first: Nola
    first-init: N.
    last: duToit
    marker: 1
  - first: Edward
    first-init: E.
    last: Mulrow
    marker: 1
footnote:
  - content: Corresponding author
    type: corresp               # set type for the corresponding author
    id: 1
affiliation:
  - institution: National Opinion Research Center (NORC)
    country: United States
    marker: 1
  - institution: Iowa State University
    department: Statistics
    country: United States
    marker: 2
bibliography: references.bib
abstract: "The use of visuals is a key component in scientific communication, and decisions about the design of a data visualization should be informed by what design elements best support the audience's ability to perceive and understand the components of the data visualization. We build on the foundations of Cleveland and McGill's work in graphical perception, employing a large, nationally-representative, probability-based panel of survey respondents to test perception in statistical charts. Our findings provide actionable guidance for data visualization practitioners to employ in their work."
keywords: 'list, keywords, here, 3-6, are, fine'
preamble: >
  \usepackage{amsfonts,amsmath,amssymb,amsthm}
  \usepackage{booktabs}
  \usepackage{graphicx}
format:
  html: default
editor: 
  markdown: 
    wrap: sentence
---

```{=latex}
  \newcommand{\hh}[1]{{\textcolor{orange}{#1}}}
  \newcommand{\kr}[1]{{\textcolor{teal}{#1}}}
  \setlength{\parindent}{0pt}
```
\hh{hello?}

@@

<!-- fig-width: 6.5 -->

<!-- fig-height: 4.5 -->

<!-- template: template-jdsart.cls.tex -->

@@

<!-- What is our main objective? -->

\kr{Should the abstract match -- or be close to -- our SDSS short abstract?}

Outline for introduction:

-   What do viewers see when we show them a data chart?

-   Visuals allow us to make comparisons between values

-   Design of a data visualization impacts viewer ability to make those comparisons

    -   And more broadly, impacts how viewers interact with a chart

-   There is prior work studying this topic

    -   Cleveland and McGill laid the foundations

    -   Mackinlay expanded to theoretical scales

    -   Heer/Bostock expanded to larger study population

-   However, prior work has utilized specific populations of study respondents.
    These populations are not representative of the general population, a common target audience for data visualization and scientific communication work

-   Our work focuses on the following:

    -   Utilizing a survey panel as a mechanism to test perception in data visualizations

    -   Testing whether prior results from convenience samples hold with a nationally representative sample

    -   Studying variations in data visualization design and their impact on respondent behavior and accuracy of participants' responses.

-   This paper presents a series of tests completed and the resulting findings.

-   The remainder of the paper is organized as follows:

    -   we first describe the stimulus design

    -   then we describe the study population (and survey design?)

    -   subsequently, we share analyses of resulting survey responses across a series of stimuli

    -   finally, we discuss implications of this study and next steps.

What do viewers see, when we show them a data chart?
A crucial step in the process of 'understanding' a chart, is that a visual allows us to make comparisons between its parts.
@clevelandGraphicalPerceptionTheory1984 defined for the purpose of their seminal study the better visual as the one that allows viewers to make more accurate comparisons.
Based on mappings of quantitative variables to different graphical elements, Cleveland and McGill's study resulted in a ranking of perceptual tasks from most accurate to least accurate.
@mackinlayAutomatingDesignGraphical1986 took this approach and expanded the tasks and their order to ordinal and nominal scales, as shown in @fig-perceptual-tasks.

@@

<!-- Bias from crowd-sourcing? sample size?  -->

Crowd workers are known to be biased towards male, young and relatively higher education: @borgoCrowdsourcingInformationVisualization2017

@@

```{r}
#| label: fig-aligned-unaligned
#| fig-cap: "The only difference between the two pairs of rectangles A, B and C, D is their alignment, i.e. A and C are of identical size, as are B and D. While the predominant response for the unaligned pair on the left is 'they are of the same size', more than half of the viewers respond with 'D is bigger' to the aligned pair of bars on the right."
#| fig-subcap:
#| - "**Unaligned bars**"
#| - "**Aligned bars**"
#| layout-nrow: 1
#| fig-align: "center"
#| out-width: 50%
knitr::include_graphics("images/round5-modified.PNG")
knitr::include_graphics("images/round3-modified.PNG")

```

## Study design -- stimulus

<!-- motivate the use of just noticeable differences -->

Asking perceptual questions in a survey is different from the controlled environment of a cognitive lab, where these kind of questions would usually be addressed.
This means, that instructions to participants have to be delivered in a very short and easily understandable, because questions arising from the task can not be answered.
Similarly, rather than asking the same (or similar) type of question with varied signals multiple hundred times, in a survey we can ask only a few questions.
In order to observe any effect, we need to ask questions that are perceptually hard, which means that we need to ask questions about stimuli that are close to our perceptual threshold.
The **Just-Noticeable Difference** (JND) is defined as the smallest difference that will be detected 50% of the time.
We are using results from studies on barcharts and pie charts [@luModelingJustNoticeable2022] to inform the differences in charts shown to survey panelists.

-   How do structural design choices in a data visualization impact viewers' ability to identify the larger of two elements?

-   How do aesthetic design choices in a data visualization impact viewers' ability to identify the larger of two elements?

@@

Aesthetic design choices (structural choices seems stronger/there could be a lot to talk about there... should we skip aesthetic on this one?)

@@

Binary accuracy (correct/incorrect -- 'they are the same' is incorrect here)

@@

I've noted this below, but: how to model?
Ordinal response?
Binary (certain or very certain vs everybody else)?

## Study design - survey population

We utilize NORC's AmeriSpeak Omnibus survey, which utilizes a probability-based panel and surveys a sample of 1,000 nationally representative adults 18 and older.
The advantage of using a probability-based approach is two-fold.
First, we have access to a large sample of survey participants and thus have greater power in making inference about graphical perceptional abilities.
Second, the sample is representative of the general adult public in the U.S., which is an important target audience for scientific communication.

@@

<!-- describe stimulus -->

Both charts in \autoref{fig-tasks-12} show the same data with slight modifications to the order of the levels -- the first and second level in each of the bars are reversed between the left and the right chart.
Participants were asked to compare the relative sizes of the tiles marked A and B (C and D, respectively) and select the correct response out of the possible choices:

@@

<!-- |                                   |                                   | -->

<!-- |-----------------------------------|-----------------------------------| -->

<!-- | **Unaligned positions**           | **Aligned positions**             | -->

<!-- | ![Unaligned positions marked A and B](../SDSS-2023/images/vertical-ab.PNG) | ![Aligned positions marked C and D](../SDSS-2023/images/vertical-cd.PNG) | -->

@@

# Results

How should we group results?

-   By stimulus/structural variation?

-   By response (binary, ordinal, certainty, time spent, etc.)

Results to discuss:

-   Differences in accuracy between aligned and unaligned bar

-   Differences in accuracy between framed, floating, and general pie

-   Differences in timing and zooming behavior across all designs

-

@@

Question for Ed: If we compare a full sample to a split sample, do we still want to weight these 'equally'?

Analysis should be done using the 'survey' package and weights should be taken into account.

@@

The data used for assessing the accuracy of comparisons in \autoref{tbl-tasks-12} is collected in two rounds of the NORC Omnibus survey.
Rounds 1 and 2 are combined by adjusting the weights with $\lambda$ = `r round(lambda[1],3)` for an effective sample size of `r round(neff(round_12$weight),1)`.
Figure \ref{fig-alignment}(a) shows that more than twice the number of responses is accurate, when the tiles are aligned along the same axis.
Because each participant was shown both versions of the chart, we can use a paired $t$-test to compare mean accuracy between the two charts.
The resulting $t$-statistic is highly significant ($t$ statistic: `r round(paired$statistic,1)`, df: `r paired$parameter`, $p$-value: \< 2.2e-16).

@@

STRUCTURAL VARIATION -- START HERE

@@

\% yes across each different structural condition

@@

\% zoomed by device + alignment (already have this chart)

@@

Model 2 (this may not be feasible for comparison depending on what level the 'TOTALTIME' is captured at):

@@

Same models as above, but I'm not sure how we want to do the response.
Ordinal response?
Binary (certain or very certain vs everybody else)?

AESTHETIC VARIATION -- ONLY IF TIME

Binary accuracy (correct/incorrect -- 'they are the same' is incorrect here) across structural choices

@@

Same models as above, but I'm not sure how we want to do the response.
Ordinal response?
Binary (certain or very certain vs everybody else)?

@@

Key findings to discuss:

-   Design choices impact viewer behavior

-   Design choices impact viewer accuracy

-   Reproduced prior convenience sample results using a large, nationally-representative survey population

-   Surveys can be used to ask these types of questions -- in the midst of other topics and with a limited number of questions, we are able to ask perception questions and produce results consistent with prior studies

-

# Supplementary Material {.unnumbered}

-   **Participant Data (Linear):** Link to csv file with the data.
-   **Data Analysis Code:** Link to an html document with annotated code chunks.
